Evaluating the Performance and Energy Efficiency of the \textsc{COSMO-ART} model system

### 2014-04-25 paper_enahpc14/Sec1:Related.tex
Manuel's changes

\begin{itemize}
\item Application centric energy-efficiency study of distributed
  multi-core and hybrid {CPU-GPU} systems \citep{Cumming-2014}
\item Stella: A domain-specific language and tool for structured grid
  methods \citep{Gysi-2014}
\item Using compiler directives to port large scientific applications
  to GPUs: An example from atmospherci science \citep{Lapillonne-2014}
\item An analysis of energy-optimized lattice-Boltzmann CFD
  simulations from the chip to the highly parallel level
  \citep{Wittmann-2013}
\item Energy efficiency vs. performance of the numerical solution of
  PDEs: An application study on a low-power ARM-based cluster
  \citep{Goddeke-2013}
\item Analysis of Computing and Energy Performance of Multicore, NUMA,
  and Manycore Platforms for an Irregular Application
  \citep{Castro-2013}
\item Exploring performance and power properties of modern multi-core
  chips via simple machine models \citep{Hager-2013}
\end{itemize}

>>>>

While the  TOP500 list was  introduced over 20  years ago to  rank the
performance of HPC systems  worldwide, it is quite recently that energy
efficiency  become  a  critical  constraint  in the  way  to  exascale
computing.   Since 2007,  the Green500  power  measurement methodology
encourages the design,  procurement and management of energy-efficient
infrastructures to contain performance in an affordable and competitive
power  envelope.   However,  the state-of-the-art  research  assessing
performance  and energy  efficiency  of applications  is still scarce.

\emph{Padoin et al.}   \cite{Padoin-2013} investigate performance and
power  consumption  of  an  agroforestry  application  and  show  that
changing workload  can drastically improve energy efficiency  of CPU +
GPU heterogeneous architectures.

\emph{Ou  Pang et  al.}   \cite{Ou-2012} compare  ARM  and Intel  x86
workstations  clusters  and   conclude  that  ARM-based  clusters  are
advantageous  with   lightweight  applications  in   terms  of  energy
efficiency.

\emph{G\"oddeke et al.}  \cite{Goddeke-2013} evaluate weak and strong
scalability of PDE solvers on a cluster of 96 ARM dual-core processors
and demonstrate  that the ARM-based  cluster can be more  efficient in
terms of energy-to-solution compared to x86-based cluster.

\emph{Wittmann  et  al.}   \cite{Wittmann-2013}  perform  a  thorough
analysis of  a lattice-Boltzmann method based CFD  simulation on Intel
Sandy Bridge  processors   and  show   extrapolated  results   on  a
petascale-class machine.

% \emph{Cumming  et  al.}   \cite{Cumming-2014}  present a  simple  and
% practical  methodology  looking at  energy  minimization  that can  be
% applied to various applications.

### 2014-04-25 paper_enahpc14/Sec2:COSMO-ART.tex
Manuel's changes

The  model  system COSMO-ART  described  in  \citep{Vogel-2009}, is  a
regional  to  continental scale  model  coupled  online  to the  COSMO
numerical weather  prediction and climate  model \citep{Baldauf-2011}.
It  incorporates  sophisticated  modules  for  gaseous  chemistry  and
aerosol dynamics  and allows the online calculation  of reactive trace
substances and their interaction  with the atmosphere.  Detailed model
description   can  be   found   in  \citep{Bangert-2012,   Knote-2011,
  Knote-2013}.

@@

Establishing effective energy performance benchmarking of a code under
intense development such as COSMO-ART is a challenging task because of
the  absolute necessity that  results must  be reproducible  within an
expected variance for the duration of the Exa2Green project. To define
a baseline,  it was necessary  to find a run-configuration  capable of
being  recreated in  all subsequent  versions.  Here  we give  a brief
overview of the model setup  for our performance and energy efficiency
evaluation.

Three-dimensional simulations are performed over large parts of Europe
and the Mediterranean~Sea for April 13th 2010 in order to benefit from
around  half  day  of  sun  exposure and  therefore  ensure  a  proper
activation of the chemistry cycle.  The calculation domain corresponds
to the CORDEX-EU-44 domain and is covered by a grid of $222\times 216$
points with  a horizontal resolution of  $0.22\,^{\circ}$, i.e., 50~km
in  both directions  and 40~vertical  layers.  These  24-hour forecast
simulations  are not  preceded  by  a spin-up  phase  to build-up  the
simulated gaseous and  aerosol concentrations.  This condition entails
that at the model initialization,  there is a period of adjustment for
the model  to reach its equilibrium  state and minimize  the effect of
initial conditions for gases and aerosols on the model predictions.

@@

The   considered  COSMO-ART   model  system   is  configured   with  a
semi-Lagrangian    horizontal   advection    scheme    with   tricubic
interpolation  and selective filling  diffusion option  in combination
with   the   dynamical    core   using   Runge-Kutta   time   stepping
\citep{COSMO-PartI-2011}.    It  also   makes  use   of   the  Kinetic
PreProcessor solver (KPP) for  the resolution of atmospheric chemistry
ordinary  differential equations \citep{Damian-2002}.   Concerning the
modelling  of  wet  deposition  in  aerosols, the  baseline  has  only
indirect  cloud feedbacks  but  does not  include in-cloud  scavenging
(rainout) and below-cloud  scavenging (washout) yet.  Amongst physical
parameterisations,   precipitation  formation   is   performed  by   a
two-moment cloud microphysics.

>>>>

The model  system \cosmoart described  in \cite{Vogel-2009},
is  a  regional to  continental  scale  model  coupled online  to  the
\textsc{Cosmo}   numerical  weather   prediction  and   climate  model
\cite{Baldauf-2011}.   It   incorporates  sophisticated  modules  for
gaseous  chemistry   and  aerosol  dynamics  and   allows  the  online
calculation of  reactive trace  substances and their  interaction with
the  atmosphere.    Detailed  model   description  can  be   found  in
\cite{Bangert-2012, Knote-2011, Knote-2013}.

@@

Establishing effective energy performance benchmarking of a code under
intense development  such as \cosmoart is  a challenging task
because of  the absolute necessity  that results must  be reproducible
within  an  expected  variance  for  the  duration  of  the  Exa2Green
pro\-je\-ct.  To   define  a  baseline,   it  was  necessary  to   find  a
run-configuration  capable  of   being  recreated  in  all  subsequent
versions.  Here  we give a brief  overview of the model  setup for our
performance and energy efficiency evaluation.

Three-dimensional simulations are performed over large parts of Europe
and the Mediterranean~Sea for April 13$^{th}$ 2010, near the equinox, in order to approximately
have a half  day  of  sun  exposure and  therefore  ensure  a typical
activation of the chemistry cycle.  The calculation domain corresponds
to the CORDEX-EU-44 domain and is covered by a grid of $222\times 216$
points with  a horizontal resolution of  $0.22\,^{\circ}$, i.e., 50~km
in  both directions  and 40~vertical  layers.  These  24-hour forecast
simulations  are not  preceded  by  a spin-up  phase  to build up  the
simulated gaseous and  aerosol concentrations.  This condition means 
that the model initialization needs a period of adjustment to reach its equilibrium  state and minimize  the effect of
initial conditions for gases and aerosols on the model predictions.

@@

The considered  \cosmoart model  system is configured  with a
semi-Lagrangian    horizontal   advection    sche\-me    with   tricubic
interpolation  and selective filling  diffusion option  in combination
with   the   dynamical    core   using   Runge-Kutta   time   stepping
\cite{COSMO-PartI-2011}.    It  also   makes  use   of   the  Kinetic
PreProcessor solver (KPP) for  the solution of atmospheric chemistry
ordinary  differential equations \cite{Damian-2002}.   Concerning the
modelling  of  wet  deposition  in  aerosols, the  baseline  has  only
indirect  cloud feedbacks  but  does not  include in-cloud  scavenging
(rainout) and below-cloud  scavenging (washout) yet.  Amongst physical
parameterizations,   precipitation  formation   is   performed  by   a
two-moment cloud microphysics.

### 2014-04-25 paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

For the experiments with \tinto platform, we also analyze the contribution of the MPI library to the energy consumption and how the use of the blocking versus polling message-passing policies can potentially render energy savings. Specifically, the OpenMPI 1.6.5 library, 
features two operation modes, blocking and polling, which can be selected before \cosmoart
is launched. In the polling configuration (the default mode) ``MPI'' continually polls
the network interface to check for the completion of an event (e.g. send or receive).
This usually yields low latency but high CPU utilization. Thus, one can expect that this mode 
attains the best performance, possibly at the cost a higher energy usage. In the blocking
mode, the CPU is yield to other processes/threads if there are no incoming messages.

\subsection{Time-power-energy analysis of \cosmoart}
\label{subsec:4.2}

\begin{figure}[htbf]
  \begin{center}
    \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Monch.eps}
    \caption{\monch: Isola E1 Rack 2 Total Power.}
    \label{fig:1}
  \end{center}
\end{figure}

\begin{figure}[htbf]
  \begin{center}
    \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Pilatus.eps}
    \caption{\pilat: Isola HD Total Power.}
    \label{fig:2}
  \end{center}
\end{figure}

Figures~\ref{fig:1}   and   \ref{fig:2}   account   respectively   for
\monch's Isola E1 Rack  2 and \pilat' Isola HD total
power measurements for  1-day or 2-day simulations. On  the Intel Ivy
Bridge EP  based cluster (i.e.  \monch),  the 1-day simulation
was issued  only twice due  to usage restrictions. As  time resolution
was set to one update every  5 minutes for power sampling, the average
power consumption was computed by considering 6 values for each single
run.  On the Intel Xeon E5 based cluster (i.e.  \pilat), the
1-day simulation  was issued  four times and  a 2-day run  only once.
Similarly, the average power consumption was computed by considering 4
values  for  each  single  1-day  run  and 9  values  for  the  2-days
run. Corresponding results are gathered in Table~\ref{tab:3}.

\begin{table}[htbf]
  \begin{center}
    \caption{Average power consumption (W) of the platforms}
    \label{tab:3}
    \begin{tabular}{cccc}
      \hline\noalign{\smallskip}
      \textbf{\scriptsize{Xeon E5}} & \textbf{\scriptsize{Ivy Bridge}} & \textbf{\scriptsize{Xeon  E5645}} & \textbf{\scriptsize{Xeon  E5645}}\\
      & & \textbf{\scriptsize{(Polling)}} & \textbf{\scriptsize{(Blocking)}} \\
      \noalign{\smallskip}\hline\noalign{\smallskip}
      18035.0 & 12622.5 & 3713.6 & 3651.8 \\ 
      \noalign{\smallskip}\hline
    \end{tabular}
  \end{center}
\end{table}

@@

Averaged TTS and ETS results over 10 repetitions of 24h simulation on \tinto are shown in Figure~\ref{fig:4}. Correponding results for average total power are shown in Table~\ref{tab:3}.  
As expected, the average total power dissipated by \cosmoart while using the blocking MPI policy is about 2\,\% less than that consumed by the polling policy. This is due to the power-friendly (blocking) waits performed by the CPU cores while using the blocking MPI mode, in contrast to the polling waits that take place when the polling MPI mode is set. From the point of view of the execution time, there is a slight reduction for the blocking MPI mode. One can expect that the costs of the interruptions used in the blocking MPI policy can lead to an increase of the total execution time, however, applications with extensive computing, as e.g. \cosmoart, may show better performance with the blocking modes. This effect works in favor with our results since the energy consumed is also reduced (2\,\%) just by changing the behavior of the OpenMPI library to the blocking policy, i.e, with the \texttt{--mca mpi\_yield\_when\_idle} parameter set in the \texttt{mpirun} command.

\begin{figure}[htbf]
  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-0.eps}
  \caption{Time-to-solution and energy-to-solution comparison between
    Xeon E5 and Ivy Bridge-EP architectures for a 24h simulation.}
  \label{fig:3}
\end{figure}

\begin{figure}[htbf]
  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-5.eps}
  \caption{Mean time-to-solution and energy-to-solution on
    \tinto for a 24h simulation using the polling and blocking MPI policies. The 95\,\% confidence intervals of Student's $t$ distribution (10 samples) clearly indicates 
    an improvement for blocking policy in both TTS and ETS.}
  \label{fig:4}
\end{figure}

\subsection{Power-performance tracing on \tinto}
\label{subsec:4.3}

A full tracing experiment is conducted on \tinto in order to capture an overall power
profile at a much finer resolution using the attached PDUs in each node. First we run \cosmoart 
for a full-day simulation and analyze the power consumption. Second we correlate performance 
with power traces of two 1-hour time frames during the day: mid-day (11h--12h) and
midnight (23h--24h). All the experiments were performed using 192 \cosmoart processes
m 12 processes in each of node.

@@

In order to obtain an initial overview of the power dissipation of the system model, we obtained power profiles leveraging our \pmlib power-measurement framework over a full-day simulation of \cosmoart. These profiles, obtained with both polling and blocking MPI policies, are depicted in  Figure~\ref{fig:9}. The first observation for the polling MPI policy is that the first and last hours of the day dissipate slightly less power than the mid-day hours. We relate these variations due to the chemical reactions and aerosols computed by ART, that are taking place during daylight hours but not during night. In this case peak power consumption is about 3821\,W. Looking at the power profile using the blocking MPI policy, we observe that this pattern is repeated, favoring the reduction of power during night hours in a more noticeable way than that for the polling case.The peak power dissipation during day hours in this mode is about 3768\,W, which means a reduction of around 50\,W with respect to the polling mode. Also, we notice systematic power drops in each hour simulated. We relate them to periods in which the processes were blocked waiting for incoming messages or synchronizations, thus yielding the CPU to low usage during. Our final observation is the reduction of the TTS in a factor of 2\,\% with the blocking MPI policy with respect to the polling one.

To gain more insights about the power-performance behavior of \cosmoart, we obtained power-per\-for\-man\-ce traces so that they allow us to correlate the type of tasks being executed with the total power dissipated at each time (see Figure~\ref{fig:11}). For this experiment we limit the simulation time to 1 hour because \emph{i}) the pattern of the computation performed by \cosmoart is repeated hourly and, \emph{ii}) the weight of a full-day simulation trace can not be easily handled using our machines and memory available (each trace file was about 45\,GB). To face the problem we shrink the simulations to 1 hour and test a time frame during midday (from 11h to 12h) and during midnight (from 23h to 24h), in which \cosmoart perform different internal computations due to the different chemical reactions that happen during day and night. At the same time we also leverage the two already stated MPI policies: polling and blocking.

All the traces in Figure~\ref{fig:11} are divided in different phases. In first phase all the processes perform group communications and execute MPI-receive primitives (e.g. \texttt{MPI\_Recv}). The second performs computation bursts, in between group communications, blocking sends, immediate receives and MPI-wait primitives are executed. Finally, group communications interlaced with computation and a significant synchronization phase is performed. From the performance point of view, one can notice a slight reduction of the TTS that blocking MPI mode leads over the polling MPI policy (around 2\,\%). A reduction of power of about  using the blocking MPI policy is produced only in routines that actually perfom synchronizations, group communications and MPI-receive/-wait primitives. Thus, the reduction of the pair power-energy directly depends on the percentage of time in which blocking MPI operations are executed. For our 1 hour simulation traces, an average of 50\,\% of the time the processes were blocked. Taking into account that  the total power reduction during blocking phases is about 5\,\%, one can compute the reduction of power, i.e., $50\,\% \times 5\,\% = 2.5\%$, which is consistent power reduction obtained for the full-day simulation (2\,\%).

\begin{figure*}[ht]
  \centering
  \hspace{0.8cm}
  \scalebox{0.55}{\input{Figs/24hour.pstex_t}}
  \caption{24 hours simulation trace using the MPI blocking mode on \tinto.}
  \label{fig:9}
\end{figure*}

>>>>

For the experiments on \tinto platform, we also analyze the contribution of the MPI library to the energy consumption and how the use of the blocking versus polling message-passing policies can potentially render energy savings. Specifically, the OpenMPI 1.6.5 library, 
features two operation modes, blocking and polling, which can be selected before \cosmoart
is launched. In the polling configuration (the default mode) ``MPI'' continually polls
the network interface to check for the completion of an event (e.g. send or receive).
This usually yields low latency but high CPU utilization. Thus, one can expect that this mode 
attains the best performance, possibly at the cost a higher energy usage. In the blocking
mode, the CPU is yield to other processes/threads if there are no incoming messages~\cite{PDP12}.

\subsection{Time-power-energy analysis}
\label{subsec:4.2}

Figures~\ref{fig:1}   and   \ref{fig:2}   account   respectively   for
\monch's Isola E1 Rack  2 and \pilat' Isola HD total
power measurements for  1-day or 2-day simulations. On  the Intel Ivy
Bridge EP  based cluster (i.e.  \monch),  the 1-day simulation
was issued  only twice due  to usage restrictions. As  time resolution
was set to one update every 5 minutes for power sampling, the averaged
power dissipation was computed by considering 6 values for each single
run.  On the Intel Xeon E5 based cluster (i.e.  \pilat), the
1-day simulation  was issued  four times and  a 2-day run  only once.
Similarly, the averaged power dissipation was computed by considering 4
values  for  each  single  1-day  run  and 9  values  for  the  2-days
run. Corresponding results are gathered in Table~\ref{tab:3}.

\begin{figure}[htbf]
  \centering
    \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Monch.eps}
    \caption{Isola E1 Rack 2 total power of \monch.}
    \label{fig:1}
\end{figure}

\begin{figure}[htbf]
  \centering
    \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Pilatus.eps}
    \caption{Isola HD total power dissipation of \pilat.}
    \label{fig:2}
\end{figure}

\begin{table}[htbf]
  \centering
    \caption{Averaged power dissipation (W) of the platforms.}
    \label{tab:3}
    \begin{tabular}{cccc}
      \hline\noalign{\smallskip}
      \textbf{\scriptsize{Xeon E5}} & \textbf{\scriptsize{Ivy Bridge}} & \textbf{\scriptsize{Xeon  E5645}} & \textbf{\scriptsize{Xeon  E5645}}\\
      & & \textbf{\scriptsize{(Polling)}} & \textbf{\scriptsize{(Blocking)}} \\
      \noalign{\smallskip}\hline\noalign{\smallskip}
      18035.0 & 12622.5 & 3713.6 & 3651.8 \\ 
      \noalign{\smallskip}\hline
    \end{tabular}
  \end{table}

@@

Averaged TTS and ETS results over 10 repetitions of 24h simulation on \tinto are shown in Figure~\ref{fig:4}. The corresponding results for the averaged total power are also shown in Table~\ref{tab:3}.  
As expected, the average total power dissipation by \cosmoart while using the blocking MPI policy is about 2\,\% less than that consumed by the polling policy. This is due to the power-friendly (blocking) waits performed by the CPU cores while using the blocking MPI mode, in contrast to the polling waits that take place when the polling MPI mode is set. From the point of view of the execution time, there is a slight reduction for the blocking MPI mode. One can expect that the costs of the interruptions used in the blocking MPI policy can lead to an increase of the execution time, however, applications with extensive computing, as e.g. \cosmoart, may show better performance with the blocking modes. This effect works in favor with our results since the energy consumed is also reduced (2\,\%) just by changing the behavior of the OpenMPI library to the blocking policy, i.e, with the \texttt{--mca mpi\_yield\_when\_idle} parameter set in the \texttt{mpirun} command.

\begin{figure}[htbf]
  \centering
  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-0.eps}
  \caption{Time-to-solution and energy-to-solution comparison between
    Xeon E5 and Ivy Bridge-EP architectures for a 24h simulation.}
  \label{fig:3}
\end{figure}

\begin{figure}[htbf]

  \centering
  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-5.eps}
  \vspace{-1cm}
  \caption{Mean time-to-solution and energy-to-solution on
    \tinto for a 24h simulation using the polling and blocking MPI policies. The 95\,\% confidence intervals of Student's $t$ distribution (10 samples) clearly indicates 
    an improvement for blocking policy in both TTS and ETS.}
  \label{fig:4}
\end{figure}

\begin{figure*}[ht]
  \centering
  \hspace{0.6cm}
  \scalebox{0.64}{\input{Figs/24hour.pstex_t}}
  \caption{24 hours simulation power trace using the polling and blocking MPI policies on \tinto.}
  \label{fig:9}
\end{figure*}

\subsection{Power-performance profiling and tracing}
\label{subsec:4.3}

A full tracing experiment is conducted on \tinto in order to capture an overall power
profile at a much finer resolution using the attached PDUs in each node. First we run \cosmoart 
for a 1-day simulation and analyze the power consumption. Second we correlate performance 
with power traces of two 1-hour time frames during the day: mid-day (11h--12h) and
midnight (23h--24h). All the experiments were performed using 192 cores using 12 MPI processes
per node on \tinto.

@@

In order to obtain an initial overview of the power dissipation of the system model, we obtained power profiles leveraging our \pmlib power-measurement framework over a 1-day simulation of \cosmoart. These profiles, produced with both polling and blocking MPI policies, are depicted in  Figure~\ref{fig:9}. The first observation for the polling MPI policy is that the first and last hours of the day dissipate slightly less power than the mid-day hours. We relate these variations due to the different chemical reactions, aerosols, radiations and clouds formation processes, computed by ART, taking place during daylight hours but not during night. In this case the peak power consumption is about 3821\,W. Looking at the power profile using the blocking MPI policy, we observe that the 1-hour pattern is repeated along the 24-hours, favoring the reduction of power during night hours in a more noticeable way than that with the polling policy. The peak power dissipation during day hours of while the blocking MPI policy is leveraged is about 3768\,W, which means a reduction of around 50\,W with respect to the polling mode. Also, we notice systematic power drops for each simulated hour. We associate them to periods in which the processes were blocked waiting for incoming messages or synchronizations, thus yielding the CPU to low usage. Our final observation is the reduction of the TTS by 2\,\% with respect to the blocking MPI policy.

To gain more insights about the power-performance behavior of \cosmoart, we obtained power-per\-for\-man\-ce traces to allow us correlate the type of tasks being executed with the total power dissipation along the time (see Figure~\ref{fig:11}). For this experiment we limit the simulation time to 1 hour because \emph{i}) the pattern of the computation performed by \cosmoart is repeated hourly and, \emph{ii}) the weight of a 1-day simulation trace can not be easily handled using our machines and memory available (each trace file was about 45\,GB). To face the problem we shrink the simulations to 1 hour and test a time frame during midday (from 11h to 12h) and  midnight (from 23h to 24h), in which \cosmoart perform different internal computations. At the same time we leverage the two already stated MPI policies: polling and blocking.

All the traces in Figure~\ref{fig:11} are divided in different computation phases. In first phase all the processes perform group communications and execute MPI-receive primitives (e.g. \texttt{MPI\_Recv}). The second one performs computation bursts, combined in between with group communications, blocking sends, immediate receives and MPI-wait primitives are executed. Finally, group communications interlaced with computation and a significant synchronization phase are performed. From the performance point of view, one can notice a reduction of 2\,\% in the TTS that blocking MPI mode leads over the polling MPI policy. On the other hand an averaged reduction of power dissipation of 2\,\% is attained with the blocking MPI policy only in routines that actually perform synchronizations, group communications and MPI-receive/-wait primitives. Thus, the reduction of the power-energy pair directly depends on the percentage of time in which blocking MPI operations are executed. For the 1 hour simulation traces, an average of 50\,\% of the time the processes were blocked. Taking into account that  the total power reduction during blocking phases is about 5\,\%, one can compute an approximation of the power savings, i.e., $50\,\% \times 5\,\% = 2.5\%$, which is consistent power reduction obtained for the full-day simulation (around 2\,\%). However, we expect higher power-energy savings if future architectures tend proportional computing~\citep{Barroso:2007:CEC:1339817.1339894}.
